Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects

### 2025-05-05 paper.qmd

---
title: "Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects"

filters:
  - acronyms
  - highlight-text
  - abstract-section

crossref:
  eq-title: Equation     # (default is "")

format:
  ieee-pdf:
    keep-tex: true
    # fig-pos: 'H'
    journaltype: conference
    papersize: a4
    include-in-header:
      - file: media/preamble.tex
    include-before-body:
      - text: |
          % NOTE: make siunitx's \qty work without clashing with the one in physics
          \let\qty\SI
  ieee-html: 
    html-math-method: mathjax
    include-in-header:
      file: media/siunitx.html

author:
  - id: es
    name: Enrico Spinielli
    affiliations:
      - ref: aiu
    orcid: 0000-0001-8584-9131
    email: enrico.spinielli@eurocontrol.int
    attributes:
      corresponding: true
  - name: Rainer Koelle
    affiliations: 
      - ref: aiu
    orcid: 0000-0003-2810-1404
    email: rainer.koelle@eurocontrol.int
  - name: Quinten Goens
    affiliations: 
      - ref: aiu
    orcid: 0000-0003-2187-1924
    email: quinten.goens@eurocontrol.int
affiliations:
  - id: aiu
    name: EUROCONTROL
    department: Aviation Intelligence Unit
    city: Brussels
    country: Belgium

bibliography: media/references.bib

acronyms:
  insert_loa: false
  fromfile:
    - ./media/acronyms.yml

---

# Abstract
<!-- teaser --> The recent decades observed an increasing number of space activities resulting in higher numbers of re-entries at the end of the mission lifespan or orbital decay. 
The risk of collision between uncontrolled re-entering space debris and aircraft
is  increasing due to both ever-growing number of airborne flights and 
of orbiting bodies (and relevant re-entries). 
<!-- why important --> National and international authorities need to balance safety and economic
impact of adopted measures should a re-entry event be announced, usually on short
notice. <!-- no hard line breaks in abstracts -->
<!-- approach / experiments --> We employ hourly and daily spatial flight densities calculated from
trajectories as recorded by EUROCONTROL's Network Manager and
binned with the Uber H3 geospatial indexing system.  
Impact risk probabilities are then calculated based on previous theoretical
work. 
<!-- results / conclusions --> This work enhances the model by employing using actual flown flight trajectories as a basis for time-varying flight densities rather than relying on proxies such as the population density or ADS-B/Mode S message counts.
These densities combined with a buffered area along the eventual path of
re-entry, can provide mid-air impact probabilities to authorities as an aid
for deciding on safety measures such as airspace closures and/or re-routings.

@@

<!-- zooming in on problem -->
With forecasts predicting a steadily growing number of flights, see @fig-forecast, and the modern
race to commercialise space, there is a need to re-evaluate the risks of
collision and casualties due to uncontrolled debris re-entries.
The increasing risk landscape poses challenges in terms of oversight and risk management.
There is a growing demand by overseeing authorities for new methodologies and tools to assess which 
safety measures are the most adequate to minimise potential casualties and economic
impact.

The increase in space operations does not exist in a vacuum. 
The continual growth of air traffic - despite temporal bumps - is evidenced by the surging number of flights in Europe [@forecastUpdate2025] and across the world. 
On-going development/deployment activities of mega-constellations of satellites in Low Earth Orbit \acr{LEO}, 
such as already active Starlink and OneWeb or planned ones (e.g. Kuiper, Starshield, Xingwang) will add thousand of orbiting objects with a short operational life. 
Some of these will inevitably re-enter the Earth's atmosphere in an uncontrolled manner, c.f. @tbl-megaconstellations [@megaconstellations].

@@

<!-- ESTABLISH SOME CONTEXT ---------------------------------------------- -->
The modelling and tracking of uncontrolled re-entries poses challenges.
Actual and historic data is sparse, and the size, shape, and attitude or rotation of re-entering objects increase the level of uncertainty for modelling.
Common approaches employ a mix of handling the uncertainty of object or debris movement, the associated exposed area, including temporal uncertainty windows.

      <!-- is there more to say and provide context? -->

<!-- APPROACH / METHOD --------------------------------------------------- -->
<!-- it would be cool, if we can add a diagram here that shows the path from data, major processing building blocks to the results -->
<!-- as a minimum, we should say something about the WHAT -->
In general, the method for estimating the risk to commercial aircraft from re-entering debris objects revolves around estimating the exposure of a single flight and the population of flights to the re-entering object.
In that respect, a robust determination of flight density is a key aspect.

@@

This work focusses on the European region.
The hourly flight density estimation is based on flown \acr{IFR} trajectories as
recorded by EUROCONTROL's \acr{NM}[^1].
To decrease spatial complexity, the data is binned using Uber's H3 geospatial indexing system [@uber-h3]. 
The reduction of the spatial complexity enables fast processing.
For this work, we employ a H3-cell-resolution of 3. 
This corresponds to hexagons of an average area $\qty{12393}{km^2}$ and an average edge length of
$\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.

@@

::: {#fig-cells}

```{r cells-map}
#| out-width: "120%"
#| echo: false
#| fig-scap: "H3 cells at resolution 3"
knitr::include_graphics("media/figures/cells_3.png")
```

H3 cells (red) at resolution 3 and their centres (blue) covering the bounding
box of 43 EUROCONTROL Member States \acr{FIR}s (blue shaded polygon).

:::

@@

<!-- next sentence feels like an orphan ... do we need this here? ------>
Next sections will develop the mathematical background for the evaluation of collision expectation of which traffic density is a key component.

@@

::: {#fig-density-summer2024-res-3}

```{r density-winter-res-3}
#| out-width: "110%"
#| width: "9cm"
#| echo: false
knitr::include_graphics("media/figures/flight_density_2024-07-05_3.png")
```

@@

## Aircraft effective exposed area{#sec-eea}

@@

$$\begin{aligned}
P &= \Delta A \times \sigma(\varphi, i) = 1000 \times \sigma\Bigl(\frac{\pi}{6}, \frac{\pi}{4}\Bigl) \\
  &= 1000 \times \num{2.490756e-15} = \num{2.490756e-12}
\end{aligned}$$

@@

::: {#fig-data-densities}

```{r densities}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/cubes.png")
```

@@

<!-- what was shown in the paper ------------------------------------------------------------>
This paper expands on previous work.
With a focus on Europe, we establish an upper bound for the traffic density based on actual flights.

@@

        WHAT ARE THINGS THAT ARE "WEAKNESSES" = STUFF THAT CAN BE IMPROVED? IT WOULD BE GOOD TO HAVE A LINE OR TWO ON THIS.

@@

<!-- application of the results -------------------------------------------------------------->
We show how National and/or International authorities could prepare for the more and more likely and frequent occurrence of uncontrolled re-entry of space objects that could pose a safety concern on civilian aviation.
The techniques we implemented on realistic data and sound algorithms can contribute to the bag of tools in the hands of technical and political entities aiming at guaranteeing the utmost safety and minimal economic disruption to civil aviation in Europe.

>>>>

---
title: "Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects"
filters:
  - acronyms
  - highlight-text
  - abstract-section
crossref:
  eq-title: Equation     # (default is "")

format:
  ieee-pdf:
    keep-tex: true
    # fig-pos: 'H'
    journaltype: conference
    papersize: a4
    include-in-header:
      - file: media/preamble.tex
    include-before-body:
      - text: |
          \let\qty\SI
          % NOTE: make siunitx's \qty work without clashing with the one in physics
  ieee-html: 
    html-math-method: mathjax
    include-in-header:
      file: media/siunitx.html
author:
  - id: es
    name: Enrico Spinielli
    affiliations:
      - ref: aiu
    orcid: 0000-0001-8584-9131
    email: enrico.spinielli@eurocontrol.int
    attributes:
      corresponding: true
  - name: Rainer Koelle
    affiliations: 
      - ref: aiu
    orcid: 0000-0003-2810-1404
    email: rainer.koelle@eurocontrol.int
  - name: Quinten Goens
    affiliations: 
      - ref: aiu
    orcid: 0000-0003-2187-1924
    email: quinten.goens@eurocontrol.int
affiliations:
  - id: aiu
    name: EUROCONTROL
    department: Aviation Intelligence Unit
    city: Brussels
    country: Belgium
bibliography: media/references.bib
acronyms:
  insert_loa: false
  fromfile:
    - ./media/acronyms.yml
editor: 
  markdown: 
    wrap: 72
---
```{r}
```

```{r}
```

# Abstract

<!-- teaser -->

The recent decades observed an increasing number of space activities
resulting in higher numbers of re-entries at the end of the mission
lifespan or orbital decay. The risk of collision between uncontrolled
re-entering space debris and aircraft is increasing due to both
ever-growing number of airborne flights and of orbiting bodies (and
relevant re-entries). <!-- why important --> National and international
authorities need to balance safety and economic impact of adopted
measures should a re-entry event be announced, usually on short notice.
<!-- no hard line breaks in abstracts -->
<!-- approach / experiments --> We employ hourly and daily spatial
flight densities calculated from trajectories as recorded by
EUROCONTROL's Network Manager and binned with the Uber H3 geospatial
indexing system.\
Impact risk probabilities are then calculated based on previous
theoretical work. <!-- results / conclusions --> This work enhances the
model by using actual flown flight trajectories as a basis for
time-varying flight densities rather than relying on proxies such as the
population density or ADS-B/Mode S message counts. These densities
combined with a buffered area along the eventual path of re-entry, can
provide mid-air impact probabilities to authorities as an aid for
deciding on safety measures such as airspace closures and/or
re-routings.

@@

<!-- zooming in on problem -->

With forecasts predicting a steadily growing number of flights, see
@fig-forecast, and the modern race to commercialise space, there is a
need to re-evaluate the risks of collision and casualties due to
uncontrolled debris re-entries. The increasing risk landscape poses
challenges in terms of oversight and risk management. There is a growing
demand by overseeing authorities for new methodologies and tools to
assess which safety measures are the most adequate to minimise potential
casualties and economic impact.

The increase in space operations does not exist in a vacuum. The
continual growth of air traffic - despite temporal bumps - is evidenced
by the surging number of flights in Europe [@forecastUpdate2025] and
across the world. On-going development/deployment activities of
mega-constellations of satellites in Low Earth Orbit \acr{LEO}, such as
already active Starlink and OneWeb or planned ones (e.g. Kuiper,
Starshield, Xingwang) will add thousand of orbiting objects with a short
operational life. Some of these will inevitably re-enter the Earth's
atmosphere in an uncontrolled manner, c.f. [@tbl-megaconstellations;
@megaconstellations].

@@

<!-- ESTABLISH SOME CONTEXT ---------------------------------------------- -->

The modelling and tracking of uncontrolled re-entries poses challenges.
Actual and historic data is sparse, and the size, shape, and attitude or
rotation of re-entering objects increase the level of uncertainty for
modelling. Common approaches employ a mix of handling the uncertainty of
object or debris movement, the associated exposed area, including
temporal uncertainty windows.

```         
  <!-- is there more to say and provide context? -->
```

<!-- APPROACH / METHOD --------------------------------------------------- -->

<!-- it would be cool, if we can add a diagram here that shows the path from data, major processing building blocks to the results -->

<!-- as a minimum, we should say something about the WHAT -->
In general, the method for estimating the risk to commercial aircraft
from reentering debris objects revolves around estimating the exposure
of a single flight and the population of flights to the re-entering
object. In that respect, a robust determination of flight density is a
key aspect.

@@

This work focusses on the European region. The hourly flight density
estimation is based on on flown \acr{IFR} trajectories as recorded by
EUROCONTROL's \acr{NM}[^1]. To decrease spatial complexity, the data is
binned using Uber's H3 geospatial indexing system [@uber-h3]. The
reduction of the spatial complexity enables fast processing. For this
work, we employ a H3-cell-resolution of 3. This corresponds to hexagons
of an average area $\qty{12393}{km^2}$ and an average edge length of
$\qty{68.979}{km}$ [@h3resolutions] as depicted in @fig-cells.

@@

::: {#fig-cells}
```{r cells-map}
#| out-width: "120%"
#| echo: false
#| fig-scap: "H3 cells at resolution 3"
knitr::include_graphics("media/figures/cells_3.png")
```

H3 cells (red) at resolution 3 and their centres (blue) covering the
bounding box of 43 EUROCONTROL Member States \acr{FIR}s (blue shaded
polygon).
:::

@@

<!-- next sentence feels like an orphan ... do we need this here? ------>
Next sections will develop the mathematical background for the
evaluation of collision and casualties risk of which traffic density is
a key component.

@@

::: {#fig-density-summer2024-res-3}
```{r density-winter-res-3}
#| out-width: "110%"
#| width: "9cm"
#| echo: false
knitr::include_graphics("media/figures/flight_density_2024-07-05_3.png")
```

@@

## Aircraft effective exposed area {#sec-eea}

@@

$$
\begin{aligned}
P &= \Delta A \times \sigma(\varphi, i) = 1000 \times \sigma\Bigl(\frac{\pi}{6}, \frac{\pi}{4}\Bigl) \\
  &= 1000 \times \num{2.490756e-15} = \num{2.490756e-12}
\end{aligned}
$$

@@

::: {#fig-data-densities}
```{r densities}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/cubes.png")
```

@@

<!-- what was shown in the paper ------------------------------------------------------------>

This paper expands on previous work. With a focus on Europe, we
establish an upper bound for the traffic density based on actual
flights.

@@

```         
    WHAT ARE THINGS THAT ARE "WEAKNESSES" = STUFF THAT CAN BE IMPROVED? IT WOULD BE GOOD TO HAVE A LINE OR TWO ON THIS.
```

@@

<!-- application of the results -------------------------------------------------------------->
